---
title: "In-Class Exercise 07"
description: |
  Lesson 7: Modelling, Visualising and Analysing Network Data with R
author:
  - name: Rakendu 
    url: https://www.linkedin.com/in/rakendu-ramesh/
    affiliation: Singapore Management University
    affiliation_url: https://scis.smu.edu.sg/
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      fig.retina = 3)
```

## Setting up

```{r}
packages = c('tidyverse', 'tidygraph', 'ggraph','visNetwork','lubridate','clock')

for(p in packages){
  if(!require(p,character.only = TRUE)){
    install.packages(p)
  }
  library(p, character.only = TRUE)
}
```

# Importing the data

```{r}
GAStech_nodes <- read_csv("data/GAStech_email_node.csv")

GAStech_edges <- read_csv("data/GAStech_email_edge-v2.csv")


```

# Data Wrangling

```{r}
GAStech_edges <- GAStech_edges %>%
  mutate(SentDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SentDate,
         label =TRUE,
         abbr = FALSE)) 

```

### Wrangling Attribute

Select all the records where the main subject is Work related. Then group by source, Target and weekday and cound the number of records and create a new column called Weight. We also filter by records where source is not same as target. and alsofilter by Weight >1

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source,target,Weekday) %>%
  summarise(Weight =n()) %>%
  filter(source != target) %>%
  filter(Weight > 1) %>%
  ungroup

```
## Building graph model

Note that the there cannot be nodes without edges between them. It is nt possible to build the graph in that case.

```{r}
gastech_graph <- tbl_graph(nodes = GAStech_nodes,
                           edges = GAStech_edges_aggregated,
                           directed = TRUE)

gastech_graph
```

```{r}
set.seed(1234)

ggraph(gastech_graph,
       layout = "fr")+
  geom_edge_link()+
  geom_node_point() +
  theme_graph()


```
## Modifying the nodes

```{r}

ggraph(gastech_graph,
       layout = "stress")+
  geom_edge_link()+
  geom_node_point(aes(colour = Department,
                      size = 3)) +
  theme_graph()
```
fr is force directed layout and will keep on changing. Hence important to set.seed so that the layout doesnt change. When the fr generate the layout, it will use gravitational force and push force.
Where as for stress(default) layout, the layout will be same without set.seed.

## Modifying the edges

```{r}
ggraph(gastech_graph,
       layout = "stress")+
  geom_edge_link(aes(width=Weight),
                 alpha =0.2)+
  scale_edge_width(range = c(0.1,5))+
  geom_node_point(aes(colour = Department,
                      size = 3)) +
  theme_graph()
```
## Working with Facet

```{r}

ggraph(gastech_graph,
       layout = "stress")+
  geom_edge_link(aes(width=Weight),
                 alpha =0.2)+
  scale_edge_width(range = c(0.1,5))+
  geom_node_point(aes(colour = Department,
                      size = 2)) +
  theme_graph() +
  facet_edges((~Weekday))
```

```{r}
ggraph(gastech_graph,
       layout = "stress")+
  geom_edge_link(aes(width=Weight),
                 alpha =0.2)+
  scale_edge_width(range = c(0.1,5))+
  geom_node_point(aes(colour = Department,
                      size = 3)) +
  theme_graph()
```
## Working with Facet by node

```{r}

ggraph(gastech_graph,
       layout = "stress")+
  geom_edge_link(aes(width=Weight),
                 alpha =0.2)+
  scale_edge_width(range = c(0.1,5))+
  geom_node_point(aes(colour = Department,
                      size = 2)) +
  theme_graph() +
  facet_nodes((~Department))
```

## How to incorporate centrality measure in our visualization

Compute the connectivity index

```{r}
gastech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness()) %>%
  ggraph(layout = "stress")+
  geom_edge_link(aes(width=Weight),
                 alpha =0.2)+
  scale_edge_width(range = c(0.1,5))+
  geom_node_point(aes(colour = Department,
                      size = betweenness_centrality)) +
  theme_graph()

```

## Interactivity for Network Graph using visNetwork

Data Preparation

```{r}

GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
    summarise(weight = n()) %>%
  filter(from!=to) %>%
  filter(weight > 1) %>%
  ungroup()
```

```{r}

visNetwork(GAStech_nodes,
           GAStech_edges_aggregated)
```
Push ensures that they are spread out and pull/gravitational force makes sure that they stay together.

```{r}

visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr")
```

```{r}
GAStech_nodes <- GAStech_nodes %>%
  rename(group = Department)
```

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout()

```