---
title: "Take-home Exercise 01"
description: |
  Creating data visualisation beyond default.
author:
  - name: Rakendu 
    url: https://www.linkedin.com/in/rakendu-ramesh/
    affiliation: Singapore Management University
    affiliation_url: https://scis.smu.edu.sg/
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

# Overview

In this Take-home exercise, we will apply the skills learnt in Visual Analytics Lesson 1 to reveal the demographics of the City of Engagement, Ohio USA by using appropriate static statistical graphics methods. We will process the data using **tidyverse** family of packages. We will then use **ggplot2** and its extensions to prepare the statistical graphics to understand the demographics of the city of Engagement, Ohio USA. 

### Dataset
The dataset used is available for download [*here*](https://vast-challenge.github.io/2022/).

# Getting Started

We will first load the required packages using the below code chunk

```{r}
packages = c('tidyverse','ggdist','gghalves')
for(p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p, character.only = T)
}

```

# Importing Data

The code chunk below imports *Exam_data.csv* from the data folder by using [`read_csv()`](https://readr.tidyverse.org/reference/read_delim.html) function of [**readr**](https://readr.tidyverse.org/index.html) package into R and save it as a tibble dataframe called *exam_data*.

```{r}
exam_data <- read_csv("data/Exam_data.csv")

```

# A Simple Bar Chart

The code chunk below creates a bar chart byusing geom_bar() of ggplot2.

```{r}

ggplot(data=exam_data,
       aes(x = RACE)) +
  geom_bar()

```


